Parallel Adaptive Importance Sampling

### 2016-06-28 PAIS.tex
Final minor corrections
Some minor corrections and typos. References cleaned up.

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e: simon.cotter@manchester.ac.uk} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and complexity, the need for efficient
  methods which can exploit the parallel architectures which are
  prevalent in high performance computing increases. In this paper, we
  aim to develop a framework for scalable parallel MCMC algorithms. At
  each iteration, an importance sampling proposal distribution is
  formed using the current states of all of the chains within an
  ensemble. Once weighted samples have been produced from this, a
  state-of-the-art resampling method is then used to create an evenly
  weighted sample ready for the next iteration. We demonstrate that
  this parallel adaptive importance sampling (PAIS) method outperforms
  naive parallelisation of serial MCMC methods using the same number
  of ensemble members, for low dimensional problems, and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, approximate multinomial resampling (AMR), which while not
  as accurate as other schemes is substantially costly for large
  ensemble sizes, which can then be used in conjunction with PAIS for
  complex problems.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
%\subsection{Lit. review}
%\subsection{Optimal transport resampler}
%\subsection{Function Space MCMC}
Markov chain Monte Carlo (MCMC) methods are a powerful family of tools
that allow us to sample from complex probability distributions. MCMC
methods were first developed in the 70s\cite{hastings1970monte}, and
with the development of faster, more
powerful computers, have become ever more important in a whole range
of fields in statistics, science and engineering.  In particular, when
considering Bayesian inverse problems, each MCMC step may involve the
numerical solution of one or more PDE. As many samples are usually
required before Monte Carlo error is reduced to acceptable levels,
these types of problem remain frustratingly out of our grasp.

Many advances have been made in the field of MCMC to design ever more
complex methods that propose moves more intelligently, leading to
rapidly converging methods. Function space versions of standard methods
such as the random walk Metropolis-Hastings (RWMH) algorithm or the
Metropolis adjusted Langevin algorithm (MALA), whose convergence rates
are independent of dimension have been
developed\cite{cotter2013mcmc}. The hybrid (or Hamiltonian) Monte
Carlo (HMC) method uses Hamiltonian dynamics in order to propose and
accept moves to states which are a long way away from the current
position\cite{sexton1992hamiltonian}, and function space analogues of
this have also been proposed\cite{beskos2011hybrid}. Riemann
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.  These
methods allow us to explore the posterior distribution more fully 
with fewer iterations.

Simultaneously, great strides are continually being made in the
development of computing hardware. Moore's law, which predicted that
the number of transistors that can to fit on a single microchip will
double every two years, has been largely followed since the early
70s\cite{moore1998cramming}. In recent times, it has become necessary
to use parallel architectures to exploit Moore's law. The efficient
exploitation of these facilities is the key to solving many of the
computational challenges that we currently face.

@@

\section{Preliminaries}\label{Sec:Prelim}
In this Section we will introduce preliminary topics and algorithms
that will be referred to throughout the paper.
\subsection{Bayesian inverse problems}
In this paper, we focus on the use of MCMC methods for characterising
posterior probability distributions in Bayesian inverse problems. We
wish to learn about a particular unknown quantity $u$, of which we are
able to make direct or indirect noisy observations. For now
we say that $x$ is a member of a Hilbert
space $X$.

@@

One advantage of the particle filter is that there are convergence
results for this method as the number of particles is increased. One
downside is that the effective sample size decreases at each
iteration, resulting in degeneration of the approximation of the
posterior. One way to tackle
this is to employ a resampling scheme. The aim of a successful
resample is to take your unevenly weighted ensemble and return a new
ensemble of particles with even weights which is highly correlated to
the original samples. Another downside is that the required ensemble
size increases quickly with dimension, making it difficult to use in
high-dimensional problems.

@@

Importance sampling can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen, from which we
can draw samples. Each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use an ensemble of states,
each coming from a particular time point in an MCMC chain, to construct a proposal
distribution which will be as close as possible to the target density. If this
ensemble is large enough, the distribution of states will be representative
of the target density.

@@

In Section \ref{Sec:Num}, we will demonstrate how the algorithm
performs, primarily using RWMH proposals. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
ensemble members can improve on the original MCMC algorithm and lead to
convergence in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set this could lead to a
large saving of computational cost. We have observed that
using more complex (and hence more expensive) kernels $\nu$, does not
significantly improve the speed of convergence of the algorithm.

@@

Efficient selection of scaling parameters in MCMC algorithms is
critical to achieving optimal mixing rates and hence achieving fast
convergence to the target density. One aspect worthy of  consideration
with the PAIS, is finding an appropriate $\nu$ such that
$\chi$ is a close approximation to the posterior density $\pi$. If the
proposal distribution is too over-dispersed, then the algorithm will often
propose states in the tails of the distribution, resulting in larger
variance of the weights, and therefore slower convergence to the
posterior distribution. Similarly, if the proposal distribution is
under-dispersed, the proposals will be highly correlated with the
previous states, and the algorithm will take a long time to fully explore
the parameter space. It is therefore necessary to find a proposal
distribution which is slightly over-dispersed to ensure the entire
posterior is explored\cite{gelman1992inference}, but is as close to the
posterior as possible.

@@

The alternative to using adaptive procedures to tune the scaling
parameters is to perform exploratory simulations to find the optimal
regimes by trial and error. This can be very costly and the optimal
parameters cannot realistically be found to more than a couple of
significant figures. It is therefore important that MCMC algorithms
provide a feasible adaptive strategy to bypass this exploratory stage.

@@

MCMC algorithms can be assessed by comparing their approximation of
the posterior to the analytic distribution, in cases where the
posterior distribution can be computed using alternative methods. To
assess this, a distance metric on distributions must be chosen. Examples are the
relative error between the sample moments and the posterior's moments, or the
relative $L^2$ error between the true density, $\pi(x|D)$, and the
constructed histogram. The relative error in the $m$-th moment is
given by:
\[
	\left|\frac{N^{-1}\sum_{i=1}^N \! x_i^m - \mathbb{E}[X^m]}{\mathbb{E}[X^m]}\right|,
\]
where $\{x_i\}_{i=1}^N$ is a sample of size $N$. The relative $L^2$
error between a continuous function to a piecewise constant function,
$e$, can be given by considering the difference in mass between the
normalised histogram of the samples and the posterior distribution
over a set of disjoint sets or ``bins'':
\begin{equation}\label{eqn:L2_error}
	e^2 = \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a - vB_i\right]^2 \Big/ \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a\right]^2,
\end{equation}
where the regions $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional
histogram bins, so that $\bigcup_i R_i \subseteq X$ and
$R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the
volume of each bin, and $B_i$ is the value of the $i$th bin. This
metric converges to the standard definition of the relative $L^2$
error as $v\rightarrow 0$.

@@

The statistic $\neff$ is easier to deal with than the variance of the
weights, as it varies between $1$ and $M$ as opposed to the variance
which can vary over many orders of magnitude.
Therefore it is preferable as a means of tuning the scaling
parameter. In the following numerics, we use the effective sample size
calculated at each iteration and averaged across the entire run to
tune the scaling parameters. We do this because when we tune this
parameter on the fly, we use the single-iteration average to estimate
optimality. The optimal scaling parameter found using the global
optimum of the variance of the weights is also included for
comparison. Note that for the majority of iterations the value of
$\neff$ is an overestimate of this statistic over a larger number of
samples. Rare events which result in a large importance weight bring
this statistic down, and care must be made not to overfit the proposal
variance to $\neff$ on an iteration by iteration basis. This can be
accounted for by increasing the variance slightly once the adaptive
algorithm has arrived on a value using single iteration values of $\neff$.

@@

From here on in, we will use the random walk proposal,
\begin{equation}y = x_{i-1}+\beta \omega_{i-1}, \qquad \omega_{i-1}
  \sim \mathcal{N}(0,1) \qquad {\rm i.i.d.}
\end{equation}
where $x_{i-1}$ is our current state and $y$ is our proposed state. In the numerics section we will refer to $\beta$ as the scaling parameter.

@@

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with $M$ for the three different
samplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

@@

In this example we compare the naively parallelised RWMH algorithm with its
PAIS variant, the PAIS-RW algorithm. The PAIS algorithm is implemented using the
ETPF to perform the resampling step. We asses the performance of the PAIS algorithm
using the relative $L^2$ error defined in Equation~\ref{eqn:L2_error}, as well as the
relative error in the first moment.

@@

Consider the simple case of a linear observation operator $\G(x) = x$, where the prior on $x$ and the observational noise follow Gaussian distributions. Then, following Equation~\ref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $x_\text{ref}=4$ noisily such that
\[
D = \mathcal{G}(x) + \eta \sim \mathcal{N}(\mathcal{G}(x_\text{ref}),\sigma^2) = \mathcal{N}(4,0.01).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
problem. A KL divergence of zero indicates that two distributions are
identical almost everywhere.

@@

{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} for one million iterations, again with $M=50$. The simulations are repeated 32 times using the optimal parameters found in (1). The performance of the algorithms is judged by the convergence of the relative $L^2$ error statistic in Equation~\ref{eqn:L2_error}.

@@

Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio compared to the error analysis and the variance of the weights. The relative $L^2$ error graph is noisy, but it is clear that the maximum in the effective sample size and the minimum in the variance of the weights are both close to the minimum in the relative $L^2$ error. Due to this we say that the estimate of the effective sample size found by averaging the statistic over each iteration is a good indicator for the optimal scaling parameter. In general this indicator underestimates the value found by using the entire sample.

@@

Both adaptive algorithms are run with initial values of $\beta=1$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and by the end of the simulation window is matching its performance. The ARWMH algorithm does not perform quite as well, this is possibly due to the fact that the acceptance rate cost function is not smooth at the optimal value making it difficult to minimise.

@@

We look at an `easy' problem, $B_1$, which has a KL divergence of
0.880, and a `harder' problem, $B_2$, which has a KL divergence of
3.647. Problem $B_1$ has two modes which are not too far apart. In
$B_2$ we increase the distance between the two modes which has the
effect of increasing the expected number of iterations that it takes
for a MCMC chain to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.

\subsubsection{Target Distribution} \label{sec:tar}

@@

The biggest issue for the Metropolis-Hastings algorithms when sampling
from a posterior such as the one in $B_2$ is that it is unlikely that
the correct ratio of chains will occur in each of the modes, and since there is no interaction between the chains, there is no way to remedy this problem. The PAIS algorithm tackles this problem with its resampling step. The algorithm uses its dynamic kernel to build up an approximation of the posterior at each iteration, and then compares this to the posterior distribution via the weights function. Any large discrepancy in the approximation will result in a large or small weight being assigned to the relevant chain, meaning the chain will either pull other chains towards it or be sucked towards a chain with a larger weight. In this way, the algorithm allows chains to `teleport' to regions of the posterior which are in need of more exploration. Figure~\ref{fig:BM2_suction} shows Problem $B_2$ with initially 1 chain in the positive mode, and 49 chains in the negative mode. It takes only a handful of iterations for the algorithm to balance out the chains into 25 chains in each mode. The chains switch modes without having to climb the energy gradient in the middle.

@@

It is often possible to isolate which reactions are occurring more
frequently (the fast reactions) and which are occurring less
frequently (the slow reactions). The quasi-steady-state assumption
(QSSA), which assumes that the fast reactions converge in distribution
on a timescale which is negligible with respect to the rate of
occurrence of the slow reactions. This assumption allows us to
approximate the dynamics of the slowly changing quantities in the
system by assuming that the fast quantities are in equillibrium with
respect to the fast reactions in isolation. This kind of model
reduction can be used to approximate the likelihood in an inverse
problem where we wish to recover the reaction parameters in the
system.

Let us consider a simple example by introducing the following simple
chemical system involving two chemical species $S_1$ and $S_2$:
\begin{equation}\label{eqn:full_system}
	\emptyset \xrightarrow{k_1} S_1 \xleftrightarrows[k_3]{k_2} S_2 \xrightarrow{k_4} \emptyset.
\end{equation}
Each arrow represents a reaction from a reactant to a product, with
some rate constant $k_i$, and where the rates of the reactions are
assumed to follow mass action kinetics. We assume that we are in a
parameter regime such that the reactions $S_1\rightarrow S_2$ and $S_2\rightarrow S_1$ occur
much much more frequently than the other reactions. Notice that both
chemical species are involved in fast reactions. However, the quantity
$\mathcal{S} = S_1 + S_2$ is conserved by both of the fast reactions,
and as such, this is the slowly changing quantity in this system.
The effective dynamics of $\mathcal{S}$ can be represented as follows.
\begin{equation}\label{eqn:QSSA_system}
	\emptyset \xrightarrow{k_1} \mathcal{S} \xrightarrow{\hat{k}_4} \emptyset.
\end{equation}
Here, the new reaction rate $\hat{k}_4$ is approximated through
application of the QSSA to be
\[
	\hat{k}_4(s) = \mathbb{E}\left[k_4S_2|\mathcal{S}=s\right] = \frac{k_2k_4\mathcal{S}}{k_2+k_3}.
\]
The value of $\mathbb{E}\left[k_4S_2|\mathcal{S}=s\right]$ is
approximated by finding the steady state of the ODE representing the fast subsystem of
reactions:
\[S_1 \xleftrightarrows[k_3]{k_2} S_2, \qquad S_1 + S_2 = s\]

If we assume that we know the rate constants $k_1$ and $k_4$, and
observe the system in Equation~\ref{eqn:QSSA_system}, we indirectly
observe the rates $k_2$ and $k_3$ through the effective rate
$\hat{k}_4$ of the degradation of $\mathcal{S}$. Our
observations are uninformative about these reaction rates, as there
are surfaces in parameter space along which the effective rate
$\hat{k}_4$ is invariant, leading to a highly ill-posed inverse
problem. Making the assumption that the errors in our observations of
the value of $\mathcal{S}$ are Gamma distributed with some variance
$\sigma^2$, this results in a long and thin posterior distribution on
$k_2$ and $k_3$. This type of problem is notoriously difficult to
sample from using standard MH algorithms, as the algorithms quickly
find a point on this manifold on which $\hat{k}_4$ is invariant, but
exploration along its length is slow.

>>>>

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e: simon.cotter@manchester.ac.uk} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and complexity, the need for efficient
  methods which can exploit the parallel architectures which are
  prevalent in high performance computing increases. In this paper, we
  aim to develop a framework for scalable parallel MCMC algorithms. At
  each iteration, an importance sampling proposal distribution is
  formed using the current states of all of the chains within an
  ensemble. Once weighted samples have been produced from this, a
  state-of-the-art resampling method is then used to create an evenly
  weighted sample ready for the next iteration. We demonstrate that
  this parallel adaptive importance sampling (PAIS) method outperforms
  naive parallelisation of serial MCMC methods using the same number
  of ensemble members, for low dimensional problems, and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, approximate multinomial resampling (AMR), which while not
  as accurate as other schemes is substantially less costly for large
  ensemble sizes, which can then be used in conjunction with PAIS for
  complex problems.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
%\subsection{Lit. review}
%\subsection{Optimal transport resampler}
%\subsection{Function Space MCMC}
Markov chain Monte Carlo (MCMC) methods are a powerful family of tools
that allow us to sample from complex probability distributions. MCMC
methods were first developed in the 70s\cite{hastings1970monte}, and
with the development of faster, more
powerful computers, have become ever more important in a whole range
of fields in statistics, science and engineering.  In particular, when
considering Bayesian inverse problems, each MCMC step may involve the
numerical solution of one or more PDE. As many samples are usually
required before Monte Carlo error is reduced to acceptable levels,
the application of MCMC methods to these types of problem remain frustratingly out of our grasp.

Many advances have been made in the field of MCMC to design ever more
complex methods that propose moves more intelligently, leading to
rapidly converging approximations. Function space versions of standard methods
such as the random walk Metropolis-Hastings (RWMH) algorithm or the
Metropolis adjusted Langevin algorithm (MALA), whose convergence rates
are independent of dimension have been
developed\cite{cotter2013mcmc}. The hybrid (or Hamiltonian) Monte
Carlo (HMC) method uses Hamiltonian dynamics in order to propose and
accept moves to states which are a long way away from the current
position\cite{sexton1992hamiltonian}, and function space analogues of
this have also been proposed\cite{beskos2011hybrid}. Riemann
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.  These
methods allow us to explore the posterior distribution more fully 
with fewer iterations.

Simultaneously, great strides are continually being made in the
development of computing hardware. Moore's law, which predicted that
the number of transistors that can fit onto a single microchip will
double every two years, has been largely followed since the early
70s\cite{moore1998cramming}. In recent times, it has become necessary
to use parallel architectures in order for this trend to continue. The efficient
exploitation of these architectures is the key to solving many of the
computational challenges that we currently face.

@@

\section{Preliminaries}\label{Sec:Prelim}
In this Section we will introduce preliminary topics and algorithms
that will be referred to throughout the paper.
\subsection{Bayesian inverse problems}
In this paper, we focus on the use of MCMC methods for characterising
posterior probability distributions arising from Bayesian inverse problems. We
wish to learn about a particular unknown quantity $u$, of which we are
able to make direct or indirect noisy observations. For now
we say that $x$ is a member of a Hilbert
space $X$.

@@

One advantage of the particle filter is that there are convergence
results for this method as the number of particles is increased.  One downside is that the required ensemble
size increases quickly with dimension, making it difficult to use in
high-dimensional problems. Another
downside is that the effective sample size decreases at each
iteration, resulting in degeneration of the approximation of the
posterior. One way to tackle
this is to employ a resampling scheme. The aim of a successful
resample is to take the unevenly weighted ensemble and return a new
ensemble of particles with even weights which is highly correlated to
the original samples.

@@

Importance sampling can be a very efficient method for sampling from a
probability distribution. A proposal density is chosen, from which we
can draw samples. Each sample is assigned a weight given by the
ratio of the target density and the proposal density at that
point. They are efficient when the proposal density is concentrated in
similar areas to the target density, and incredibly inefficient when
this is not the case. The aim of the PAIS is to use an ensemble of states,
each coming from a MCMC chain, to construct a proposal
distribution which will be as close as possible to the target density. If this
ensemble is large enough, the distribution of states will be representative
of the target density.

@@

In Section \ref{Sec:Num}, we will demonstrate how the algorithm
performs, primarily using RWMH proposals. We do not claim that this choice is optimal, but
is simply chosen as an example to show that sharing information across
ensemble members can improve on the original MCMC algorithm and lead to
convergence in fewer evaluations of $\G$. This is important since if
the inverse problem being tackled involves computing
the likelihood from a very large data set this could lead to a
large saving of computational cost. We have observed that
using more complex (and hence more expensive) kernels $\nu$, does not
significantly improve the speed of convergence of the algorithm for
simple examples\cite{Paul}.

@@

Efficient selection of scaling parameters in MCMC algorithms is
critical to achieving optimal mixing rates and hence achieving fast
convergence to the target density. One aspect worthy of  consideration
with the PAIS, is finding an appropriate proposal kernel $\nu$ such
that the mixture distribution
$\chi$ is a close approximation to the posterior density $\pi$. If the
proposal distribution is too over-dispersed, then the algorithm will often
propose states in the tails of the distribution, resulting in larger
variance of the weights, and therefore slower convergence to the
posterior distribution. Similarly, if the proposal distribution is
under-dispersed, the proposals will be highly correlated with the
previous states, and the algorithm will take a long time to fully explore
the parameter space, and worse, will lead regularly to states proposed
in the tails of the proposal distribution with very large weights. It is therefore necessary to find a proposal
distribution which is slightly over-dispersed to ensure the entire
posterior is explored\cite{gelman1992inference}, but is as close to the
posterior as possible.

@@

The alternative to using adaptive procedures to tune the scaling
parameters is to perform exploratory simulations to find the optimal
regimes by trial and error. This can be very costly and the optimal
parameters cannot realistically be found to more than a couple of
significant figures. It is therefore important that MCMC algorithms
provide a feasible adaptive strategy.

@@

MCMC algorithms can be assessed by comparing their approximation of
the posterior to the analytic distribution, in cases where the
posterior distribution can be computed using alternative methods. To
assess this, a distance metric on distributions must be chosen. Examples are the
relative error between the sample moments and the posterior's moments, or the
relative $L^2$ error between the true density, $\pi(x|D)$, and the
constructed histogram. The relative error in the $m$-th moment is
given by:
\begin{equation}\label{eq:34567}
	\left|\frac{N^{-1}\sum_{i=1}^N \! x_i^m - \mathbb{E}[X^m]}{\mathbb{E}[X^m]}\right|,
\end{equation}
where $\{x_i\}_{i=1}^N$ is a sample of size $N$. The relative $L^2$
error between a continuous function to a piecewise constant function,
$e$, can be given by considering the difference in mass between the
normalised histogram of the samples and the posterior distribution
over a set of disjoint sets or ``bins'':
\begin{equation}\label{eqn:L2_error}
	e^2 = \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a - vB_i\right]^2 \Big/ \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(a|D) \, \mbox{d}a\right]^2,
\end{equation}
where the regions $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional
histogram bins, so that $\bigcup_i R_i \subseteq X$ and
$R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the
volume of each bin, and $B_i$ is the value of the $i$th bin. This
metric converges to the standard definition of the relative $L^2$
error as $v\rightarrow 0$.

@@

The statistic $\neff$ is easier to deal with than the variance of the
weights, as it varies between $1$ and $M$ as opposed to the variance
which can vary over many orders of magnitude.
Therefore it is preferable as a means of tuning the scaling
parameter. In all of the numerics which follow, we use the effective sample size
calculated at each iteration and averaged across the entire run to
tune the scaling parameters. We do this because when we tune this
parameter on the fly, we use the single-iteration average to estimate
optimality. The optimal scaling parameter found using the global
optimum of the variance of the weights is also included for
comparison. Note that for the majority of iterations the value of
$\neff$ is an overestimate of this statistic over a larger number of
samples. Rare events which result in a large importance weight bring
this statistic down, and care must be made not to overfit the proposal
variance to $\neff$ on an iteration by iteration basis. This can be
accounted for by increasing the variance slightly once the adaptive
algorithm has arrived on a value using single iteration values of $\neff$.

@@

From here on in, we will use the random walk proposal,
\begin{equation}y = x_{i-1}+\beta \omega_{i-1}, \qquad \omega_{i-1}
  \sim \mathcal{N}(0,\Sigma) \qquad {\rm i.i.d.}
\end{equation}
where $x_{i-1}$ is our current state, $y$ is our proposed state and
$\Sigma$ is a covariance operator, which could come from the prior distribution. In the numerics section we will refer to $\beta$ as the scaling parameter.

@@

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with ensemble size $M$ for the three different
samplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

@@

In this example we compare the naively parallelised RWMH algorithm with its
PAIS variant, the PAIS-RW algorithm. The PAIS algorithm is implemented using the
ETPF to perform the resampling step. We assess the performance of the PAIS algorithm
using the relative $L^2$ error defined in \eqref{eqn:L2_error}, as well as the
relative error in the first moment \eqref{eq:34567}.

@@

Consider the simple case of a linear observation operator $\G(x) = x$, where the prior on $x$ and the observational noise follow Gaussian distributions. Then, following \eqref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $x_\text{ref}=4$ noisily such that
\[
D = \mathcal{G}(x) + \eta \sim \mathcal{N}(\mathcal{G}(x_\text{ref}),\sigma^2) = \mathcal{N}(4,0.01).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
problem. A KL divergence of zero indicates that two distributions are
identical almost everywhere.

@@

{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the algorithms in Section~\ref{sec:problem 1} for one million iterations, again with $M=50$. The simulations are repeated 32 times using the optimal parameters found in (1). The performance of the algorithms is judged by the convergence of the relative $L^2$ error statistic in \eqref{eqn:L2_error}.

@@

Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio
compared to the error analysis and the variance of the weights. The
relative $L^2$ error graph is noisy, but it is clear that the maximum
in the effective sample size and the minimum in the variance of the
weights are both close to the minimum in the relative $L^2$ error. Due
to this we say that the estimate of the effective sample size found by
averaging the statistic over each iteration is a good indicator for
the optimal scaling parameter. In general this indicator overestimates
the value of $\neff$ found by using the entire sample.

@@

Both adaptive algorithms are run with initial values of
$\beta=1$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and
by the end of the simulation window is matching its performance. The
ARWMH algorithm does not perform quite as well, this is possibly due
to the fact that the acceptance rate cost function is not particularly
smooth at the optimal value making it difficult to minimise.

@@

\subsubsection{Target Distribution} \label{sec:tar}
We look at an `easy' problem, $B_1$, which has a KL divergence of
0.880, and a `harder' problem, $B_2$, which has a KL divergence of
3.647. Problem $B_1$ has two modes which are not too far apart. In
$B_2$ we increase the distance between the two modes which has the
effect of increasing the expected number of iterations that it takes
for a MCMC chain to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.

@@

The biggest issue for the Metropolis-Hastings algorithms when sampling
from a posterior such as the one in $B_2$ is that it is unlikely that
the correct ratio of chains will be maintained in each of the modes, and since there is no interaction between the chains, there is no way to remedy this problem. The PAIS algorithm tackles this problem with its resampling step. The algorithm uses its dynamic kernel to build up an approximation of the posterior at each iteration, and then compares this to the posterior distribution via the weights function. Any large discrepancy in the approximation will result in a large or small weight being assigned to the relevant chain, meaning the chain will either pull other chains towards it or be sucked towards a chain with a larger weight. In this way, the algorithm allows chains to `teleport' to regions of the posterior which are in need of more exploration. Figure~\ref{fig:BM2_suction} shows Problem $B_2$ with initially 1 chain in the positive mode, and 49 chains in the negative mode. It takes only a handful of iterations for the algorithm to balance out the chains into 25 chains in each mode. The chains switch modes without having to climb the energy gradient in the middle.

@@

It is often possible to isolate which reactions are occurring more
frequently (the fast reactions) and which are occurring less
frequently (the slow reactions). The quasi-steady-state assumption
(QSSA) is the assumption that the fast reactions converge in distribution
on a timescale which is negligible with respect to the rate of
occurrence of the slow reactions. This assumption allows us to
approximate the dynamics of the slowly changing quantities in the
system by assuming that the fast quantities are in equillibrium with
respect to the fast reactions in isolation. This kind of model
reduction can be used to approximate the likelihood in an inverse
problem where we wish to recover the reaction parameters in the
system.

Let us consider a simple example by introducing the following simple
chemical system involving two chemical species $S_1$ and $S_2$:
\begin{equation}\label{eqn:full_system}
	\emptyset \xrightarrow{k_1} S_1 \xleftrightarrows[k_3]{k_2} S_2 \xrightarrow{k_4} \emptyset.
\end{equation}
Each arrow represents a reaction from a reactant to a product, with
some rate constant $k_i$, and where the rates of the reactions are
assumed to follow mass action kinetics. We denote the concentration of
species $S_i$ by $X_i$. We assume that we are in a
parameter regime such that the reactions $S_1\rightarrow S_2$ and $S_2\rightarrow S_1$ occur
much much more frequently than the other reactions. Notice that both
chemical species are involved in fast reactions. However, the quantity
$\mathcal{S} = X_1 + X_2$ is conserved by both of the fast reactions,
and as such, this is the slowly changing quantity in this system.
The effective dynamics of $\mathcal{S}$ can be represented as follows.
\begin{equation}\label{eqn:QSSA_system}
	\emptyset \xrightarrow{k_1} \mathcal{S} \xrightarrow{\hat{k}_4} \emptyset.
\end{equation}
Here, the new reaction rate $\hat{k}_4$ is approximated through
application of the QSSA to be
\[
	\hat{k}_4(s) = \mathbb{E}\left[k_4X_2|\mathcal{S}=s\right] = \frac{k_2k_4\mathcal{S}}{k_2+k_3}.
\]
The value of $\mathbb{E}\left[k_4X_2|\mathcal{S}=s\right]$ is
approximated by finding the steady state of the ODE representing the fast subsystem of
reactions:
\[S_1 \xleftrightarrows[k_3]{k_2} S_2, \qquad X_1 + X_2 = s\]

If we assume that we know the rate constants $k_1$ and $k_4$, and
observe the system in \eqref{eqn:QSSA_system}, we indirectly
observe the rates $k_2$ and $k_3$ through the effective rate
$\hat{k}_4$ of the degradation of $\mathcal{S}$. Our
observations are uninformative about these reaction rates, as there
are surfaces in parameter space along which the effective rate
$\hat{k}_4$ is invariant, leading to a highly ill-posed inverse
problem. Making the assumption that the errors in our observations of
the value of $\mathcal{S}$ are Gamma distributed with some variance
$\sigma^2$, this results in a long and thin posterior distribution on
$k_2$ and $k_3$. This type of problem is notoriously difficult to
sample from using standard MH algorithms, as the algorithms quickly
find a point on this manifold on which $\hat{k}_4$ is invariant, but
exploration along its length is slow.
